---
title: 'Geog6300: Lab 6'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Regression

```{r setup, message=FALSE, warning=FALSE}
#install.packages(car)
library(sf)
library(tidyverse)
library(tmap)
library(car)
```

**Overview:**
This lab focuses on regression techniques. You'll be analyzing the association of various physical and climatological characteristics in Australia with observations of several animals recorded on the citizen science app iNaturalist.

###Data and research questions###

Let's import the dataset.

```{r}
lab6_data<-st_read("data/aus_climate_inat.gpkg")

lab6_data<-st_as_sf(lab6_data,coords=c(x="longitud",y="latitude"),crs=4326)
```

The dataset for this lab is a 1 decimal degree hexagon grid that has aggregate statistics for a number of variables:

* ndvi: NDVI/vegetation index values from Landsat data (via Google Earth Engine). These values range from -1 to 1, with higher values indicating more vegetation.
* maxtemp_00/20_med: Median maximum temperature (C) in 2000 or 2020 (data from SILO/Queensland government)
* mintemp_00/20_med: Median minimum temperature (C) in 2020 or 2020 (data from SILO/Queensland government)
* rain_00/20_sum: Total rainfall (mm) in 2000 or 2020 (data from SILO/Queensland government)
* pop_00/20: Total population in 2000 or 2020 (data from NASA's Gridded Population of the World)
* water_00/20_pct: Percentage of land covered by water at some point during the year in 2000 or 2020
* elev_med: Median elevation (meters) (data from the Shuttle Radar Topography Mission/NASA)

There are also observation counts from iNaturalist for several distinctively Australian animal species: the central bearded dragon, the common emu, the red kangaroo, the agile wallaby, the laughing kookaburra, the wombat, the koala, and the platypus.

Our primary research question is how the climatological/physical variables in our dataset are predictive of the NDVI value. We will build models for 2020 as well as the change from 2000 to 2020. The second is referred to as a "first difference" model and can sometimes be more useful for identifying causal mechanisms.

###Part 1: Analysis of 2020 data###

We will start by looking at data for 2020. 

**Question 1** _Create histograms for NDVI, max temp., min temp., rain, and population, and water in 2020 as well as elevation. Based on these graphs, assess the normality of these variables._

```{r}
hist(lab6_data$ndvi_20_med)
ggplot(lab6_data,aes(sample=ndvi_20_med))+
  stat_qq()+stat_qq_line()

hist(lab6_data$maxtemp_20_med)
ggplot(lab6_data,aes(sample=maxtemp_20_med))+
  stat_qq()+stat_qq_line()

hist(lab6_data$mintemp_20_med)
ggplot(lab6_data,aes(sample=mintemp_20_med))+
  stat_qq()+stat_qq_line()

hist(lab6_data$rain_20_sum)
ggplot(lab6_data,aes(sample=rain_20_sum))+
  stat_qq()+stat_qq_line()

hist(lab6_data$pop_20)
ggplot(lab6_data,aes(sample=pop_20))+
  stat_qq()+stat_qq_line()

hist(lab6_data$water_20_pct)
ggplot(lab6_data,aes(sample=water_20_pct))+
  stat_qq()+stat_qq_line()

hist(lab6_data$elev_med)
ggplot(lab6_data,aes(sample=elev_med))+
  stat_qq()+stat_qq_line()
```

NDVI had a right skewed histogram and did not display normality. The max temperature histogram was left skewed and the qq line looked more normal than the NDVI line. Minimum temperature had a someone uniform histogram, slightly skewed left. It's qq line was somewhat normal as well. Rain was significantly skewed right. Population and water were extremely skewed right. Elevation was somewhat normal, slightly skewed right.

**Question 2** _Use tmap to map these same variables using Jenks natural breaks as the classification method. For an extra challenge, use `tmap_arrange` to plot all maps in a single figure._

```{r}
map1 <- tm_shape(lab6_data) +
  tm_polygons("ndvi_20_med",size=.2,style = "jenks")

map2 <- tm_shape(lab6_data) +
  tm_polygons("maxtemp_20_med",size=.2,style = "jenks")

map3 <- tm_shape(lab6_data) +
  tm_polygons("mintemp_20_med",size=.2,style = "jenks")

map4 <- tm_shape(lab6_data) +
  tm_polygons("rain_20_sum",size=.2,style = "jenks")

map5 <- tm_shape(lab6_data) +
  tm_polygons("pop_20",size=.2,style = "jenks")

map6 <- tm_shape(lab6_data) +
  tm_polygons("water_20_pct",size=.2,style = "jenks")

map7 <- tm_shape(lab6_data) +
  tm_polygons("elev_med",size=.2,style = "jenks")

tmap_arrange(map1, map2, map3, map4, map5, map6, map7)
```


**Question 3** _Based on the maps from question 3, summarise major patterns you see in the spatial distribution of these data from any of your variables of interest. How do they appear to be associated with the NDVI variable?_

Overall, temperature seemed to increase as you go north. Rainfall was higher on coastal regions, as is the population. Elevation was higher in the southeast and western regions. The largest regions covered in water were more central and near the coast, mainly in the east. The NDVI variable seemed to follow a similar pattern to increased rainfall, near the coastal regions and especially in the southeast. This was also in the coolest part of the country. The vegetation does not seem to be associated with the other variables as much.

**Question 4** _Create univariate models for each of the variables listed in question 1, with NDVI in 2020 as the dependent variable. Print a summary of each model. Write a summary of those results that indicates the direction, magnitude, and significance for each model coefficient._

```{r}
model1<-lm(ndvi_20_med~maxtemp_20_med,data=lab6_data)
summary(model1)

model2<-lm(ndvi_20_med~mintemp_20_med,data=lab6_data)
summary(model2)

model3<-lm(ndvi_20_med~rain_20_sum,data=lab6_data)
summary(model3)

model4<-lm(ndvi_20_med~pop_20,data=lab6_data)
summary(model4)

model5<-lm(ndvi_20_med~water_20_pct,data=lab6_data)
summary(model5)

model6<-lm(ndvi_20_med~elev_med,data=lab6_data)
summary(model6)
```

For every one unit increase of median maximum temperature 2020, NDVI decreases by 0.013, which is somewhat high. The p-value for the slope indicates the relationship is highly significant. The R-squared value is somewhat low (0.21), meaning NDVI cannot be heavily explained by maximum temp. 
For every one unit increase of median minimum temperature 2020, NDVI decreases by 0.012. This is very similar to the maximum temperature variable, so it is likely these two variables have similar trends with NDVI. It was statistically significant and the R-squared value was also low (0.14).
For every one unit increase in rainfall 2020, NDVI increases by 9.124e-07, which is pretty small. This value was statistically significant and the R-squared value was somewhat high (0.43), therefore NDVI can be somewhat explained by rainfall. 
For every one unit increase in population, NDVI increases by 1.500e-06, which is very small. The value is statistically significant. The R-squared value is somewhat low (0.12) which means we cannot say that NDVI is explained by population.
For every one unit increase of percentage of land covered by water, NDVI decreases by 0.18, which is a somewhat high increase. The p-value level is high, so there is insufficient evidence that these two variables are associated. 
For every one unit increase of elevation, NDVI increases by 1.787e-04. The relationship is highly significant although R-squared value is somewhat low, meaning we cannot say that NDVI can be explained by elevation.

**Question 5** _Create a multivariate regression model with the variables of interest, choosing EITHER max or min temperature (but not both) You may also choose to leave out any variables that were insignificant in Q4. Use the univariate models as your guide. Call the results._

```{r}
model_all<-lm(ndvi_20_med~maxtemp_20_med + rain_20_sum + pop_20 + elev_med, data = lab6_data)
summary(model_all)
```

**Question 6** _Summarize the results of the multivariate model. What are the direction, magnitude, and significance of each coefficient? How did it change from the univariate models you created in Q4 (if at all)? What do the R2 and F-statistic values tell you about overall model fit?_

The variables follow a similar pattern to the univariate models in terms of direction. The magnitude for the variables decreased a little. They are still statistically significant, except the population variable rised some. The R-squared value increased to 0.64, which means the data is more reliable and NDVI is pretty well explained by the variables. The F-statistic value is highly significant so the model can explain the variation in NDVI. 

**Question 7** _Use a histogram and a map to assess the normality of residuals and any spatial autocorrelation. Summarise any notable patterns that you see._

```{r}
residuals <- residuals(model_all)
lab6_data$residuals<-residuals(model_all)
hist(residuals)
tm_shape(lab6_data) +
  tm_dots("residuals", size=.2)
```

The histogram displays a somewhat normal distribution in a specific region, it looks like there is some values further to the left so the data is skewed left. The residuals displayed on the map show a couple strong negative values in the southeast and north east region. There are also some positive values in the southeast. Overall, the residuals are normal with some slightly negative ones clustered throughout. The negative residuals mean the model overestimates the predictions while the positive ones mean the model underestimates. The residuals seem to be present in more urban areas, meaning this affects how the other variables act. Overall, the negative residuals are clustered but are also located nearby the positive residuals which are clustered. 

**Question 8** _Assess any issues with multicollinearity or heteroskedastity in this model using the techniques shown in class. Run the appropriate tests and explain what their results show you._

```{r}
vif(model_all)
library(lmtest)
bptest(model_all)
```

The values in the model are all below 5, indicating there is no multicollinearity or correlation between two variables. The model indicates there is heteroskedastity since the p-value is very small. The BP value is quite large, meaning there is stronger heteroskedastity.

**Question 9** _How would you summarise the results of this model in a sentence or two? In addition, looking at the full model and your diagnostics, do you feel this is a model that provides meaningful results? Explain your answer._

The model is reliable since it shows no multicollinearity, as all the values are below 5. The variables are all statistically significant so the model provides meaningful results. The model shows strong heteroskedastity, so we cannot say the predictions are fully reliable. Overall, we can conclude a higher NDVI was primarily associated with higher rainfall. Lower NDVI was also associated with higher temperatures.

**Disclosure of assistance:** _Besides class materials, what other sources of assistance did you use while completing this lab? These can include input from classmates, relevant material identified through web searches (e.g., Stack Overflow), or assistance from ChatGPT or other AI tools. How did these sources support your own learning in completing this lab?_

I used Google to help me undertstand the meaning of multicollinearity. I also used ChatGPT to figure out how to take the residuals from the data to create the histogram.

**Lab reflection:** _How do you feel about the work you did on this lab? Was it easy, moderate, or hard? What were the biggest things you learned by completing it?_

The lab was moderately difficult, I learned more about how to read and interpret the summary from the models.


**Challenge question**

#Option 1
Create a first difference model. To do that, subtract the values in 2000 from the values in 2020 for each variable for which that is appropriate. Then create a new model similar to the one you created in question 5, but using these new variables showing the *change in values* over time. Call the results of the model, and interpret the results in the same ways you did above. Also chart and map the residuals to assess model error. Finally, write a short section that summarises what, if anything, this model tells you. 

```{r}
maxtemp_dif <- lab6_data$maxtemp_20_med - lab6_data$maxtemp_00_med
mintemp_dif <- lab6_data$mintemp_20_med - lab6_data$mintemp_00_med
rain_dif <- lab6_data$rain_20_sum - lab6_data$rain_00_sum
ndvi_dif <- lab6_data$ndvi_20_med -  lab6_data$ndvi_00_med
water_dif <- lab6_data$water_20_pct - lab6_data$water_00_pct
pop_dif <- lab6_data$pop_20 - lab6_data$pop_00

model_all2 <- lm(ndvi_dif ~ maxtemp_dif + mintemp_dif + rain_dif + water_dif + pop_dif, data = lab6_data)
summary(model_all2)

residuals2 <- residuals(model_all2)
hist(residuals2)
```

Rainfall difference has a positive relationship with NDVI difference meaning that as rainfall increases over time, NDVI increases as well. Minimum temperature difference has a negative relationship with NDVI difference meaning that as the minimum temperature increases over time, NDVI decreases. The water, population, and maximum temperature differences all are not statistically significant, meaning we cannot make any conclusions based on the values. Changes in minimum temperature and rainfall both were statistically significant, meaning they are the only reliable variables. The R-squared value is somewhat small so the model only explains 10.61% of variation of the NDVI difference. The model is not reliable to detect variability. The F-statistic value, however, is statistically significant.


#Option 2
The animal data included in this dataset is an example of count data, and usually we would use a Poisson or similar model for that purpose. Let's try it with regular OLS regression though. Create two regression models to assess how the counts of two different animals (say, koalas and emus) are associated with at least three of the environmental/climatological variables given above. Be sure to use the same independent variables in each model. Interpret the results of each model and then explain the importance of any differences in the model coefficients between them, focusing on direction, magnitude, and significance.

```{r}

```

